---
title: "LABORATORY EXAM"
author: "Paculba, Laika Jean B."
date: "2023-06-29"
output: 
  pdf_document: 
  lineheight: 5
  fontsize: 8
---

# 1 The Fractional Distillation Data

The purity of oxygen produced by a fractional distillation process is thought to be related
to the percentage of hydrocarbons in the main condensor of the processing unit. Twenty
samples are shown below.

```{r}
library(pander)

data1 <- data.frame(
  observations = 1:20,
  Purity = c(86.91, 89.85, 90.28, 86.34, 92.58, 87.33, 86.29, 91.86, 95.61, 89.86, 96.73, 99.42, 98.66, 96.07, 93.65, 87.31, 95.00, 96.85, 85.20, 90.56),
  Hydrocarbon = c(1.02, 1.11, 1.43, 1.11, 1.01, 0.95, 1.11, 0.87, 1.43, 1.02, 1.46, 1.55, 1.55, 1.55, 1.40, 1.15, 1.01, 0.99, 0.95, 0.98),
  stringsAsFactors = FALSE
)
data1$Hydrocarbon <- sprintf("%.2f", data1$Hydrocarbon)
data1$Purity <- sprintf("%.2f", data1$Purity)

pander::pander(data1)
```


###a. Create a scatter diagram for the data.

```{r}
library(ggplot2)

data1 <- data.frame(
  observations = 1:20,
  Purity = c(86.91, 89.85, 90.28, 86.34, 92.58, 87.33, 86.29, 91.86, 95.61, 89.86, 96.73, 99.42, 98.66, 96.07, 93.65, 87.31, 95.00, 96.85, 85.20, 90.56),
  Hydrocarbon = c(1.02, 1.11, 1.43, 1.11, 1.01, 0.95, 1.11, 0.87, 1.43, 1.02, 1.46, 1.55, 1.55, 1.55, 1.40, 1.15, 1.01, 0.99, 0.95, 0.98))

Purity = c(86.91, 89.85, 90.28, 86.34, 92.58, 87.33, 86.29, 91.86, 95.61, 89.86, 96.73, 99.42, 98.66, 96.07, 93.65, 87.31, 95.00, 96.85, 85.20, 90.56)
Hydrocarbon = c(1.02, 1.11, 1.43, 1.11, 1.01, 0.95, 1.11, 0.87, 1.43, 1.02, 1.46, 1.55, 1.55, 1.55, 1.40, 1.15, 1.01, 0.99, 0.95, 0.98)
  
ggplot(data = NULL, aes(x = Hydrocarbon, y= Purity)) + 
  geom_point() + geom_smooth(method = "lm", se = FALSE, color = "pink") + 
  labs(title = "Scatter Diagram of The Fractional Distillation Data",  x = "HYDROCARBON", y = "PURITY")

```
###b. The least-squares fit is .

```{r}
model <- lm(data = data1,
formula = Purity ~ Hydrocarbon)
names(model)
```
```{r}
model$coefficients
```
```{r}
model_summary <- summary(model)
model_summary$sigma
```

```{r}
# Therefore, the least square fit is y_hat = 77.46239 + (12.02816)x 
# view the fitted values
fitted.values <- fitted(model)
fitted.values
```

###c. The estimate of $\sigma^2$is.

```{r}
sigma_hat_squared <- ((model_summary$sigma)^2)
print(paste("Therefore the estimate of sigma squared is: ", sigma_hat_squared))
```


###d. Test for significance of regression in the regression model.
```{r, message=FALSE, fig.align='center', fig.height=3, fig.width=5}
library(mosaic)
model_summary$coefficients["Hydrocarbon",]
mosaic::xqt(0.975, 18)
```

###e. Use an analysis-of-variance approach to test significance of regression.
Null Hypothesis: $H_0$: $\beta_1 = 0$
Alternative Hypothesis: $H_1$: $\beta_1 != 0$ 
```{r}
t_value <- model_summary$coefficients["Hydrocarbon","t value"]
print(t_value)
```
```{r}
# Find the critical value
df <- 20
significance_level <- 0.05
critical_value <- qt(1-significance_level/2, df-2)
print(paste("The critical value is:", critical_value))
```
When $\alpha=0.05$, the critical value of t is $t_{0.025, 18} = 2.101$ and the T Value is 3.39

Since |t value| > critical value

Thus, we would reject $H_0: beta_1 = 0$.

Hence, there is a linear relationship between shear strength and the age of the propellant.
```{r}
model_summary
```



```{r, fig.align='center', fig.width=5, fig.height=3}
model_summary$fstatistic
mosaic::xqf(0.99,1,18)
```
```{r}
anova_result <- anova(model)
print(anova_result)
```
Extracting the F value
```{r}
f_value <- anova_result$'F value'[1]
print(f_value)
```
Finding the critical values
```{r}
# Find the critical value
df1 <- 1
df2 <- 20
significance_level <- 0.05
critical_fvalue <- qf(1-significance_level, df1, df2-2)
print(paste("The Critical F value is:", critical_fvalue))
```

```{r}
result <- t.test(Hydrocarbon, Purity)
p_value <- result$p.value
print(p_value)
```
The P value for this test is 0.003.

Using F test in testing the analysis of variance and choosing the significance level $\alpha=0.05$, the critical value of f is computed as $f_{0.05,1,18} = 4.41$

The F value is 11.4658 and the critical F value is 4.41. Since the F value is greater than the critical value, Hence, we reject the null hypothesis.

Therefore, there is a linear relationship between shear strength and the age of the propellant.

###f. Find a 95%CI on the slope.

```{r}
# Find a 95%CI on the slope
# Input data
y <- c(86.91, 89.85, 90.28, 86.34, 92.58, 87.33, 86.29, 91.86, 95.61, 89.86,
96.73, 99.42, 98.66, 96.07, 93.65, 87.31, 95.00, 96.85, 85.20, 90.56)
x <- c(1.02, 1.11, 1.43, 1.11, 1.01, 0.95, 1.11, 0.87, 1.43, 1.02,1.46, 1.55,
1.55, 1.55, 1.40, 1.15, 1.01, 0.99, 0.95, 0.98)
# Perform linear regression
model <- lm(y ~ x)
# Calculate 95% confidence interval
conf_interval <- confint(model, level = 0.95)
# Display the confidence interval
conf_interval
```


###g. Find a 95%CI on the mean purity when the hydrocarbon percentage is 1.00.

```{r}
# Find a 95%CI on the mean y when x percentage is 1.00.
# Input data
y <- c(86.91, 89.85, 90.28, 86.34, 92.58, 87.33, 86.29, 91.86, 95.61, 89.86,
96.73, 99.42, 98.66, 96.07, 93.65, 87.31, 95.00, 96.85, 85.20, 90.56)
x <- c(1.02, 1.11, 1.43, 1.11, 1.01, 0.95, 1.11, 0.87, 1.43, 1.02,1.46, 1.55,
1.55, 1.55, 1.40, 1.15, 1.01, 0.99, 0.95, 0.98)
# Perform linear regression
model <- lm(y ~ x)
# Set the x value for prediction

3

x_pred <- 1.00
# Predict the mean y for x_pred
y_pred <- predict(model, newdata = data.frame(x = x_pred), interval = "confidence", level = 0.95)
# Display the confidence interval
y_ci <- y_pred[, c("lwr", "upr")]
y_ci
```


# 2. The Steam Consumption Data
The number of pounds of steam used per month at a plant is thought to be related to the
average monthly ambient temperature. The past year’s usages and temperatures follow.




```{r}
library(pander)

data2 <- data.frame(
  observations = 1:12,
  Usage = c(185.79, 214.47, 288.03, 424.84, 454.68, 539.03, 621.55, 675.06, 562.03, 452.93, 369.95, 273.98),
  Temperature = c(21, 24, 32, 47, 50, 59, 68, 74, 62, 50, 41, 30),
  stringsAsFactors = FALSE
)
data2$Temperature <- sprintf("%.2f", data2$Temperature)
data2$Usage <- sprintf("%.2f", data2$Usage)

pander::pander(data2)

```

###a. Create a scatter diagram for the data.

```{r}
library(ggplot2)

data2 <- data.frame(
  observations = 1:12,
  Usage = c(185.79, 214.47, 288.03, 424.84, 454.68, 539.03, 621.55, 675.06, 562.03, 452.93, 369.95, 273.98),
  Temperature = c(21, 24, 32, 47, 50, 59, 68, 74, 62, 50, 41, 30))

Usage = c(185.79, 214.47, 288.03, 424.84, 454.68, 539.03, 621.55, 675.06, 562.03, 452.93, 369.95, 273.98)
Temperature = c(21, 24, 32, 47, 50, 59, 68, 74, 62, 50, 41, 30)
  
ggplot(data = NULL, aes(x = Temperature, y= Usage)) + 
  geom_point() + geom_smooth(method = "lm", se = FALSE, color = "pink") + 
  labs(title = "Scatter Diagram of The Fractional Distillation Data",  x = "TEMPERATURE", y = "USAGE")

```

###b. The least-squares fit is .
```{r}
model2 <- lm(data = data2,
formula = Usage ~ Temperature)
names(model2)
```
```{r}
model2$coefficients
```
```{r}
modelsummary <- summary(model2)
modelsummary$sigma
```

```{r}
modelsummary
```


```{r}
# Therefore, the least square fit is y_hat = 77.46239 + (12.02816)x 
# view the fitted values
fitted.values2 <- fitted(model2)
fitted.values2
```

###c. The estimate of $\sigma^2$

```{r}
sigma_hat_squared2 <- ((modelsummary$sigma)^2)
print(paste("Therefore the estimate of sigma squared is: ", sigma_hat_squared2))
```

###d. Test for significance of regression in the regression model.
```{r, message=FALSE, fig.align='center', fig.height=3, fig.width=5}
library(mosaic)
modelsummary$coefficients["Temperature",]
mosaic::xqt(0.975, 18)
```

###e. Use an analysis-of-variance approach to test significance of regression.
```{r}
t_value2 <- modelsummary$coefficients["Temperature","t value"]
print(t_value2)
```

```{r}
# Find the critical value
df <- 20
significance_level <- 0.05
critical_value <- qt(1-significance_level/2, df-2)
print(paste("The critical value is:", critical_value))
```
When $\alpha=0.05$, the critical value of t is $t_{0.025, 18} = 2.101$ and the T Value is 272.255

Since |t value| > critical value

Thus, we would reject $H_0: beta_1 = 0$.

Hence, there is a linear relationship between shear strength and the age of the propellant.
```{r}
modelsummary
```



```{r, fig.align='center', fig.width=5, fig.height=3}
modelsummary$fstatistic
mosaic::xqf(0.99,1,18)
```
```{r}
anova_result2 <- anova(model2)
print(anova_result2)
```

Extracting the F value
```{r}
f_value2 <- anova_result2$'F value'[1]
print(f_value2)
```

Finding the critical values
```{r}
# Find the critical value
df1 <- 1
df2 <- 20
significance_level2 <- 0.05
critical_fvalue2 <- qf(1-significance_level, df1, df2-2)
print(paste("The Critical F value is:", critical_fvalue2))
```

```{r}
result <- t.test(Temperature, Usage)
p_value2 <- result$p.value
print(p_value2)
```
The P value for this test is 0.003.

Using F test in testing the analysis of variance and choosing the significance level $\alpha=0.05$, the critical value of f is computed as $f_{0.05,1,18} = 4.41$

The F value is 74122.78 and the critical F value is 4.41. Since the F value is greater than the critical value, Hence, we reject the null hypothesis.

Therefore, there is a linear relationship between shear strength and the age of the propellant.

###f. Find a 99%CI on the slope.
```{r}
# Find a 95%CI on the slope
# Input data
y2 <- c(185.79, 214.47, 288.03, 424.84, 454.68, 539.03, 621.55, 675.06, 562.03, 452.93, 369.95, 273.98)
x2 <- c(21, 24, 32, 47, 50, 59, 68, 74, 62, 50, 41, 30)
# Perform linear regression
model2 <- lm(y2 ~ x2)
# Calculate 95% confidence interval
conf_interval2 <- confint(model2, level = 0.99)
# Display the confidence interval
conf_interval2
```

###g. Construct a 99% prediction interval on steam usage in a month with average ambient temperature of 58◦

```{r, warning=FALSE}
# Find a 95%CI on the mean y when x percentage is 1.00.
# Input data
y2 <- c(185.79, 214.47, 288.03, 424.84, 454.68, 539.03, 621.55, 675.06, 562.03, 452.93, 369.95, 273.98)
x2 <- c(21, 24, 32, 47, 50, 59, 68, 74, 62, 50, 41, 30)
# Perform linear regression
model2 <- lm(y2 ~ x2)
# Set the x value for prediction

3

x_pred2 <- 58
# Predict the mean y for x_pred2
y_pred2 <- predict(model2, newdata = data.frame(x = x_pred2), interval = "confidence", level = 0.99)
# Display the confidence interval
y_ci2 <- y_pred2[, c("lwr", "upr")]
y_ci2
```