
---
title: "LABORATORY ACTIVITY"
author: "Paculba, Laika Jean B."
date: "2023-06-02"
output: 
  pdf_document: 
  lineheight: 5
  fontsize: 8
---

# 1. The Rocket Propellant Data

A rocket motor is manufactured by bonding an igniter propellant and a sustainer propellant together
inside a metal housing. The shear strength of the bond between the two types of propellant is an
important quality characteristic. It is suspected that shear strength is related to the age in weeks
of the batch of sustainer propellant. Twenty observations on shear strength and the age of the
corresponding batch of propellant have been collected.


```{r, echo=FALSE}

library(pander)

rocketpropellant <- data.frame(
  observations = c(1:20),
  ShearStrength = c(2158.70, 1678.15, 2316.00, 2061.30, 2207.50, 1708.30, 1784.70, 2575.00, 2357.90, 2256.70, 2165.20, 2399.55, 1779.80, 2336.75, 1765.30, 2053.50, 2414.40, 2200.50, 2654.20, 1753.70),
  AgeofPropellant = c(15.50, 23.75, 8.00, 17.00, 5.50, 19.00, 24.00, 2.5, 7.50, 11.00, 13.00, 3.75, 25.00, 9.75, 22.00, 18.00, 6.00, 12.50, 2.00, 21.50),
  stringsAsFactors = FALSE
  
)
rocketpropellant$AgeofPropellant <- sprintf("%.2f", rocketpropellant$AgeofPropellant)
rocketpropellant$ShearStrength <- sprintf("%.2f", rocketpropellant$ShearStrength)

pander::pander(rocketpropellant)

```


a. Scatter diagram for the data.
```{r echo=FALSE, warning=FALSE}
library(ggplot2)

  rocketpropell <- data.frame(observations = c(1:20),
  ShearStrength = c(2158.70, 1678.15, 2316.00, 2061.30, 2207.50, 1708.30, 1784.70, 2575.00, 2357.90, 2256.70, 2165.20, 2399.55, 1779.80, 2336.75, 1765.30, 2053.50, 2414.40, 2200.50, 2654.20, 1753.70),
  AgeofPropellant = c(15.50, 23.75, 8.00, 17.00, 5.50, 19.00, 24.00, 2.5, 7.50, 11.00, 13.00, 3.75, 25.00, 9.75, 22.00, 18.00, 6.00, 12.50, 2.00, 21.50))
  
    ShearStrength = c(2158.70, 1678.15, 2316.00, 2061.30, 2207.50, 1708.30, 1784.70, 2575.00, 2357.90, 2256.70, 2165.20, 2399.55, 1779.80, 2336.75, 1765.30, 2053.50, 2414.40, 2200.50, 2654.20, 1753.70)
  AgeofPropellant = c(15.50, 23.75, 8.00, 17.00, 5.50, 19.00, 24.00, 2.5, 7.50, 11.00, 13.00, 3.75, 25.00, 9.75, 22.00, 18.00, 6.00, 12.50, 2.00, 21.50)
ggplot(data = NULL, aes(x = AgeofPropellant, y= ShearStrength)) + 
  geom_point() + geom_smooth(method = "lm", se = FALSE, color = "pink") + 
  labs(title = "Scatter Diagram of Rocket Propellant Data",  x = " AGE OF PROPELLANT", y = "Shear Strength")
```

***

## 2. Least-Squares Estimation of the Parameters

```{r, echo=TRUE, warning=FALSE}
#Use the lm() function to calculate the linear model based on the data set.
# calculate model
model <- lm(data = rocketpropell,
formula = ShearStrength ~ AgeofPropellant)
```

The model object is a list of a number of different pieces of information, which can be seen by looking
at the names of the objects in the list.

```{r}
# view the names of the objects in the model
names(model)
```

## a. The least-squares fit is
The summary() function is a useful way to gather critical information in your model.

```{r}
model$coefficients
```

```{r}
model_summary <- summary(model)
model_summary$sigma
```
```{r}
# Therefore, the least square fit is y_hat = 2627.822 + (-37 15)x or y_hat= 2627.822 - 37.15x
# view the fitted values
fitted.values <- fitted(model)
fitted.values

```

## b. The estimate of *$\sigma^2$*

```{r, out.height=50}
sigma_hat_squared <- ((model_summary$sigma)^2)
print(paste("Therefore the estimate of sigma squared is: ", sigma_hat_squared))
```

***

# 3.  Hypothesis Testing on the Slope and Intercept

```{r, message=FALSE, fig.align='center', fig.height=3, fig.width=5}
library(mosaic)
model_summary$coefficients["AgeofPropellant",]
mosaic::xqt(0.975, 18)
```

Null Hypothesis: $H_0$: $\beta_1 = 0$
Alternative Hypothesis: $H_1$: $\beta_1 != 0$ 

## a. Test for significance of regression in the rocket propellant regression model

```{r}
t_value <- model_summary$coefficients["AgeofPropellant","t value"]
print(t_value)

```

```{r}
# Find the critical value
df <- 20
significance_level <- 0.05
critical_value <- qt(1-significance_level/2, df-2)
print(paste("The critical value is:", critical_value))
```
When $\alpha=0.05$, the critical value of t is $t_{0.025, 18} = 2.101$.

Since |t value| > critical value

Thus, we would reject $H_0: beta_1 = 0$.

Hence, there is a linear relationship between shear strength and the age of the propellant.

```{r}
model_summary

```

```{r, fig.align='center', fig.width=5, fig.height=3}
model_summary$fstatistic
mosaic::xqf(0.99,1,18)
```

## b. Use an analysis-of-variance approach to test significance of regression

```{r}
anova_result <- anova(model)
print(anova_result)
```

Extracting the f-value
```{r}
f_value <- anova_result$'F value'[1]
print(f_value)
```

Finding the critical values
```{r}
# Find the critical value
df1 <- 1
df2 <- 20
significance_level <- 0.05
critical_fvalue <- qf(1-significance_level, df1, df2-2)
print(paste("The Critical F value is:", critical_fvalue))
```

```{r}
result <- t.test(AgeofPropellant, ShearStrength)
p_value <- result$p.value
print(p_value)
```
The P value for this test is 1.64e-10.

Using F test in testing the analysis of variance and choosing the significance level $\alpha=0.05$, the critical value of f is computed as $f_{0.05,1,18} = 4.41$

The F value is 163.3768 and the critical F value is 4.41. Since the F value is greater than the critical value, Hence, we reject the null hypothesis.

Therefore, there is a linear relationship between shear strength and the age of the propellant.

