---
title: ""
author: "Laika Jean B. Paculba"
date: "March 29, 2023"
ouput: pdf_document
---

List elements can be nay data type and any dimensions 
Each element can be given a name

Arrays - are n-dimensional data structures 

Heterogenous structure 
List are one- dimnesionl
data is not one time only

my_list <- c(6, TRUE, "hello")
str(my_list)
my_list <- list(6, TRUE, "hello")
str(my_list)

vector- homogenous
```
new_list <- list(
scalar = c("Hello", "Goodbye"),
matrix = matrix(1:4, nrow=2, ncol=2)
)
str(new_list)
new_list$matrix
```

Named list elements can be accessed by using $

Dataframes most used data structure, named list of vectirs 
similar to database table

two vectors of different types but same length

names <- c("John", "Ayesha")
ages <- c(31, 24)
(df <- data.frame(names, ages)) #create dataframe
str(df) #get two types of columns
dim(df) #get dimensions

```
names <- c("John", "Ayesha")
ages <- c(31, 24)
(df <- data.frame(names, ages)) 
str(df) 
dim(df) 
```

dataframe most commnonly used by data analyst 

to work with data in R, usually need to pull it in from an outside source into a dataframe
```
R data sets
```

r facilitates numerous ways of importing, databases

The read.csv 

url of data set
url <- 
# load the data set and store it as a dataframe called salespeople
salespeople <- read.csv(url)

```
url <- "https://raw.githubusercontent.com/msuiitdmsgabriel/datasets-regression/main/salespeople.csv"
salespeople <- read.csv(url)
```

read downloaded file
```
salespeople_local <- read.csv("salespeople.csv")
```

hundred of rowas, so view first few
```
dim(salespeople)
head(salespeople)
```

view a specic view by using $ sign and use 

```
salespeople$sales
salespeople$sales[6]
```

use [row, column] index to get a specific entry in the dataframe

salespeople[34,4]
salespeople[34,] #access the whole row
```
salespeople[34,4]
salespeople[34,]
salespeople[,4]
```
look at datatypes using str
```
str(salespeople)
```

see statistical summary of each column by using summary()

```{r}
summary(salespeople)
```

noote that there are missing data in this dataframe(NA)

missing data denoted as NA

function is.na() look at all values in a vector or dataframe and return TRUE, False

by adding sum() functionn, counts NA as 1

counts how many missing data in dataframe
```
is.na(salespeople$sales)
is.na(salespeople)
sum(is.na(salespeople))
```

complete.cases identify rows without NAs


```{r}
salespeople <- salespeople[complete.cases(salespeople),] #override the content of the first object with missing values
complete.cases(salespeople)
sum(is.na(salespeople)) #confirm no NAs
salespeople[complete.cases(salespeople),] #only the complete cases
```

unique() function 
extract different values -- unique(salespeople$performance)
```
unique(salespeople$performance)
```

as.numeric() -- change to numeric
as.factor()

produce factor --as.factor(salespeople$performance)
overide salespeople as factor--- salespeople$performance <- as.factor(salespeople$performance)

```
as.factor(salespeople$performance)
salespeople$performance <- as.factor(salespeople$performance)
str(salespeople)
```

# Manipulating data frames


``` 
(sales_720 <- subset(salespeople, subset = sales == 720))
```
## Inequality

select all rows that the value is greather than or equal to 700 show the first 6 rows
exclude rows that has 720
```{r}
high_sales <- subset(salespeople, subset = sales >= 700)
subset(salespeople, subset = sales != 700)
head(high_sales)
```


select specific columns using select argument
```{r}
salespeople_sales_perf <- subset(salespeople, select = c("sales", "performance"))
head(salespeople_sales_perf)
```

two dataframes with thje same name can be combined 

```{r}
low_sales <- subset(salespeople, subset = sales < 400)
#bind the rows of low_sales and high_sales togther
low_and_high_sales = rbind(low_sales, high_sales) #combined low and high sales together
head(low_and_high_sales)
```
```{r}
#two dataframes with columns each
sales_perf <- subset(salespeople, select = c("sales", "performance"))
prom_custrate <- subset(salespeople, select = c("promoted", "customer_rate"))
#bind the columns to create a dataframe with four columns
full_df <- cbind(sales_perf, prom_custrate)
head(full_df)
```

cbind makes data apad

rbind is below

## Functions and

head()
subset()
rbind()
- combine row wise

exist to perform common operattions

head()
displays the 6 rows with one required the name of the dataframe
default value n= 6

```{r}
head(full_df, n=10)
```
```{r}
?head
head(x = full_df, n = 10)
head(full_df, n10)
```
```{r}
#see the headspeople 
head(salespeople,3) #3 is used to overide the default value 6
```
```{r}
#if you dont know the right orderof the name your arguments and you can put them in any order
head(n = 3, x = salespeople)
```

# Help Functions
```
help(head)
?head
```

open help browser and do the search there

help:
examples
required and optional arguments
data type of input and output

## Writing yopur own functions
funxctions are not limited in package in R
write own functions hekpl=ful i  their 
Dont Repeat yourself(DRY)
write the same code numerous times for practice

generates  a report on a dataframe
```{r}
#create df_report function
df_report <- function(df) {
  paste("This dataframe contains", nrow(df),
  "rows and", ncol(df), 
  "columns. There are", sum(is.na(df)), 
  "NA entries"
  )
}
```
```{r}
df_report(salespeople)
df_report(sales_720)
df_report(low_and_high_sales)
df_report(salespeople_local)
```


